Non-coding RNAs of bird genomes

### 2013-08-27 paper/manuscript.tex
Added edits by Jana.

% Load packages
%\usepackage{hyperref}
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage{xspace}
%\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}

@@

\newcommand {\pg}[1]{\textcolor{blue}{#1}}
\newcommand {\ppg}[1]{\textcolor{blue}{#1}}

@@

\author{
Paul P. Gardner\correspondingauthor$^{1,2}$
\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Sarah W. Burge$^3$
\email{sb30@sanger.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E. Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Härtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\begin{abstract}
        % Do not use inserted blank lines (ie \\) until main body of text.
        %% \paragraph*{Background:} Text for this section of the abstract.       
        %% \paragraph*{Results:} Text for this section of the abstract \ldots
        %% \paragraph*{Conclusions:} Text for this section of the abstract \ldots

Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase. Although we show
extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and
more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds,
we also demonstrate apparent ``losses'' in several RNA families. These
include the divergence of some classical ncRNAs and the loss of
several snoRNAs and microRNAs. In contrast, we show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds. These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

%% 1 School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 2 Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 3 European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK. 4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom. 5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Härtelstrasse 16-18, D-04107 Leipzig, Germany

@@

\section*{Results}

In the following we explore the conservation patterns of the major
classes of ncRNA in further detail. The collection of ncRNA sequences
is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. Consequently the bulk of the
families discussed below are either conserved across the vertebrates
or have been apparently lost or diversified in the avian
lineage(s).

@@

Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
in the maturation of other functional RNAs \cite{Gardner:2010}. The
bulk of the characterised snoRNAs guide either methylation or
pseudouridylation modifications, primarily of rRNAs but also
spliceosomal RNAs. The two types of modifications are guided by two
different types of RNA, the C/D and the H/ACA box snoRNAs respectively,
each with a characteristic cohort of motifs and secondary structures
\cite{Marz:2011}.

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.

Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 locii of the ancestral
vertebrate genome. We investigated these losses further.

The first cluster is found at chr11:62620797-62622484 on the human
genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
human-yeast conserved snoRNAs. These snoRNAs are located in the
inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
alligator and turtle genomes within a 3-4 KB locus, yet these have
largely been lost in the birds. Notable exceptions are five of the
eight are located in the tinamou genome, these are located on the same
scaffold and are within 2 KB of each other. Implying that SNHG1 is
conserved in the tinamou. Three of the eight are located in the
ostrich and zebrafinch genomes, again within 2 KB of each other on the
same scaffolds. This complex pattern of loss could be attributed to
many different models, e.g. multiple losses in birds, poor homology
modelling or incomplete genome sequences.

@@

The second cluster is located at chr19:49993222-49994231 on the human
genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
within a 1 KB genomic region. The turtle and alligator genomes retain
the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
the same scaffold. Within the bird genomes, the crow and rifleman each
retain a single SNORD33 and SNORD34 gene on the same scaffold. While
the ground-finch and zebrafinch retain a single SNORD33 and the bald
eagle and seriema retain a single SNORD34 (See Figure~\ref{fig:3}).
In human these snoRNAs are intronic to the host gene, ribosomal
protein L13a (RPL13A). Based on BLASTP (version 2.2.18) homology
searches for the RPL13A gene, the protein is conserved in the human
and turtle genomes and in the bald eagle, crow, rifleman and
zebrafinch avian genomes (data not shown). Therefore the conservation
patterns of the RPL13A host gene and corresponding intronic snoRNAs
largely mimic each other, each independently supports a pattern of
loss of the RPL13A gene and the intronic snoRNAs that it hosts in the
bird genomes.

\subsection*{MicroRNAs}

MicroRNAs are an important class of non-coding RNA. They have been
found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
\cite{Hinas:2007,Avesson:2012}, Viridiplantae
\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}.  The
miRNAs have been shown to regulate the expression of large numbers of
messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
22 nucleotides long which is usually processed from a larger RNA that
is characterised by a stable hairpin-shaped secondary structure.

@@

{\bf MAY BE MOVED INTO THE lncRNA PAPER!}

@@

The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
remarkably well conserved. In the human genome the corresponding
RNA-domains are located at chr7:27135743-27245922 (hg19) and are
modelled by 15 RNA-domains in total, the % (5, 4 and 6 respectively)
lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
and turtle genomes five and six of the HOX RNA-domains are conserved,
each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
the HOX cluster lncRNA RNA-domains were predicted on the same
scaffold, this gene order was preserved.

%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

@@

%(neighbor of BRCA1 gene 2) 
%MHM results!!!!

%% Another of the functions associated with lncRNAs is chromosomal dosage
%% compensation. In mammals this is carried out by the lncRNAs Xist and
%% Tsix \cite{}. In birds it has been proposed that the lncRNA MHM might
%% play a related role \cite{}.

@@

\subsection*{Pseudogenes}

The human genome is littered with pseudogenised copies of ncRNAs. The
Dfam database of human repeats estimates that approximately $54\%$ of
the human genome is composed of repeat elements
\cite{Wheeler:2013}. Of these annotations XXX\% are derived from
ncRNAs, including the large Alu family of repeats, which is derived
from a retransposed SRP RNA pseudogene \cite{}.

tRNA, RSV RNA, 
 \cite{International_Chicken_Genome_Sequencing_Consortium:2004}

\emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB

@@

@@

Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad.

%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%
%% ---------------------------------
%% BioMedCentral bibtex .BST file will be used to
%% create a .BBL file which includes the BMC XML.
%% Note that the displayed Bibliography will not be 
%% exactly as specified in the online Instructions for Authors

@@

%% \begin{figure}[ht]
%%   \includegraphics[width=0.65\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
%%   \caption[]{.}\label{fig:2}
%% \end{figure}

\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
    the left.}\label{fig:3}
\end{figure}

>>>>

% Load packages
%\usepackage{hyperref}
\usepackage{cite} % Make references as [1-4], not [1,2,3,4]
\usepackage{url} % Formatting web addresses
\usepackage{ifthen} % Conditional
\usepackage{multicol} %Columns
\usepackage{xspace}
\usepackage[utf8]{inputenc} %unicode support
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\urlstyle{rm}
\usepackage[OT1]{fontenc}

@@

\newcommand {\pg}[1]{\textcolor{blue}{#1}}
\newcommand {\ppg}[1]{\textcolor{blue}{#1}}
\newcommand {\new}[1]{\textcolor{green}{#1}}

@@

\author{
Paul P. Gardner\correspondingauthor$^{1,2}$
\email{Paul P. Gardner\correspondingauthor - paul.gardner@canterbury.ac.nz},
Jana Hertel$^5$
\email{Jana Hertel\correspondingauthor - jana@bioinf.uni-leipzig.de},
Sarah W. Burge$^3$
\email{sb30@sanger.ac.uk},
Maria Ninova$^4$
\email{Maria.Ninova@postgrad.manchester.ac.uk},
Stephanie Kehr$^5$
\email{steffi@bierdepot.bioinf.uni-leipzig.de},
Mario Fasold$^5$
\email{mario@bierdepot.bioinf.uni-leipzig.de},
Tammy E. Steeves$^1$
\email{tammy.steeves@canterbury.ac.nz},
Sam Griffiths-Jones$^4$
\email{sam.griffiths-jones@manchester.ac.uk}
and
Peter Stadler\correspondingauthor$^5$
\email{Peter Stadler\correspondingauthor - studla@bioinf.uni-leipzig.de}
}
\address{
\iid(1) School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(2) Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand.
\iid(3) European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK.
\iid(4) Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom.
\iid(5) Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, H{\"a}rtelstrasse 16-18, D-04107 Leipzig, Germany
}

@@

\begin{abstract}
Here we present the results of a large-scale bioinformatic annotation
of non-coding RNAs in 48 avian genomes. Our approach uses
probabilistic models of hand-curated families from the Rfam database
to infer conserved RNA families within each avian genome. We
supplement these annotations with predictions from the tRNA annotation
tool, tRNAscan-SE and microRNAs from miRBase. Although we show
extensive conservation of classical ncRNAs (e.g., tRNAs and rRNAs) and
more recently discovered ncRNAs (e.g., snoRNAs and miRNAs) in birds,
we also demonstrate apparent ``losses'' in several RNA families. These
include the divergence of some classical ncRNAs and the loss of
several snoRNAs and microRNAs. In contrast, we show a significant
number of lncRNAs are surprisingly well conserved between birds and
mammals including several intriguing cases where the reported
mammalian lncRNA function is not conserved in birds. These combined
results illustrate the utility of applying homology based methods for
annotating vertebrate genomes and illustrate many complex evolutionary
patterns within the avian ncRNA cohort.
\end{abstract}

@@

%% 1 School of Biological Sciences, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 2 Biomolecular Interaction Centre, University of Canterbury, Private Bag 4800, Christchurch, New Zealand. 3 European Molecular Biology Laboratory, European Bioinformatics Institute, Hinxton, Cambridge, CB10 1SD, UK. 4Faculty of Life Sciences, University of Manchester, Manchester, United Kingdom. 5 Bioinformatics Group, Department of Computer Science; and Interdisciplinary Center for Bioinformatics, University of Leipzig, Hartelstrasse 16-18, D-04107 Leipzig, Germany

@@

\section*{Results}
In the following we explore the conservation patterns of the major
classes of ncRNA in further detail. The collection of ncRNA sequences
is generally biased towards model organisms
\cite{Gardner:2010,Hoeppner:2012}. Consequently the bulk of the
families discussed below are either conserved across the vertebrates
or have been apparently lost or diversified in the avian
lineage(s).

@@

Small nucleolar RNAs (snoRNAs) are important ncRNAs that participate
in the maturation of other functional RNAs \cite{Gardner:2010}. The
bulk of the characterised snoRNAs guide either methylation or
pseudouridylation modifications, primarily of rRNAs but also
spliceosomal RNAs. The two types of modifications are guided by two
different types of RNA, the box C/D and the H/ACA snoRNAs
respectively, each with a characteristic cohort of motifs and
secondary structures \cite{Marz:2011}.

There are 66 ribosomal modification sites, guided by 59 snoRNA
families, that are preserved between \emph{H. sapiens} and
\emph{S. cerevisiae} \cite{Lestrade:2006}. Of these, 45 snoRNA
families are conserved in the bird dataset.
Over a third of the apparent losses of the yeast-human conserved
snoRNA families appear to cluster on 2 locii of the ancestral
vertebrate genome. We investigated these losses further.

The first cluster is found at chr11:62620797-62622484 on the human
genome (hg19) and contains SNORD27, SNORD29 and SNORD31 of the
human-yeast conserved snoRNAs. These snoRNAs are located in the
inside-out gene SNHG1 which hosts a total of eight C/D box snoRNAs:
SNORD25, SNORD26, SNORD27, SNORD28, SNORD29, SNORD22, SNORD30 and
SNORD31 \cite{Tycowski:1996}. Each of which are also found in the
alligator and turtle genomes within a 3-4 KB locus, yet these have
largely been lost in the birds. Notable exceptions are five of the
eight are located in the tinamou genome, these are located on the same
scaffold and are within 2 KB of each other. Implying that SNHG1 is
conserved in the tinamou. Loci with four of the eight snoRNAs can be
found in zebrafinch, ground-finch, and bald eagle. Still, three of the
eight are located in the ostrich, crow, and cockoo genomes, again
within 2 KB of each other on the same scaffolds. This complex pattern
of loss could be attributed to many different models, e.g. multiple
losses in birds, poor homology modelling or incomplete genome
sequences.

@@

The second cluster is located at chr19:49993222-49994231 on the human
genome (hg19) and contains two copies of SNORD33 and one SNORD34 all
within a 1 KB genomic region. The turtle and alligator genomes retain
the two copies of SNORD33 yet don't have an obvious SNORD34 gene on
the same scaffold. Within the bird genomes, the crow and rifleman each
retain a single SNORD33 and SNORD34 gene on the same scaffold. While
the ground-finch and bald eagle retain a single SNORD33 and the
zebrafinch and seriema retain a single SNORD34 (See
Figure~\ref{fig:3}).  In human these snoRNAs are intronic to the host
gene, ribosomal protein L13a (RPL13A). Based on BLASTP (version
2.2.18) homology searches for the RPL13A gene, the protein is
conserved in the human and turtle genomes and in the bald eagle, crow,
rifleman and zebrafinch avian genomes (data not shown). Therefore the
conservation patterns of the RPL13A host gene and corresponding
intronic snoRNAs largely mimic each other, each independently supports
a pattern of loss of the RPL13A gene and the intronic snoRNAs that it
hosts in the bird genomes.

\subsection*{MicroRNAs}
MicroRNAs are an important class of non-coding RNA. They have been
found in the genomes of Chromalveolata \cite{Cock:2010,Huang:2011},
Metazoa \cite{Lee:1993,Lau:2001,Hertel:2006}, Mycetozoa
\cite{Hinas:2007,Avesson:2012}, Viridiplantae
\cite{Reinhart:2002,Fattash:2007,Axtell:2007,Molnar:2007} and Viruses
\cite{Pfeffer:2004,Ouellet:2008,Pfeffer:2005,Landgraf:2007}. The
miRNAs have been shown to regulate the expression of large numbers of
messenger RNAs \cite{Lim:2005}. The mature miRNA product is generally
22 nucleotides long which is usually processed from a larger RNA that
is characterised by a stable hairpin-shaped secondary structure.

@@

%%{\bf MAY BE MOVED INTO THE lncRNA PAPER!}

@@

The HOX cluster lncRNAs, HOTAIRM1, HOTTIP and HOXA11-AS1, are
remarkably well conserved. In the human genome the corresponding
RNA-domains are located at chr7:27135743-27245922 (hg19) and are
modelled by 15 RNA-domains in total, the % (5, 4 and 6 respectively)
lncRNAs are ordered HOTAIRM1 (5 RNA-domains) $\rightarrow$ HOXA11-AS1
(6 RNA-domains) $\rightarrow$ HOTTIP (4 RNA-domains). In the alligator
and turtle genomes five and six of the HOX RNA-domains are conserved,
each supporting an ancestral gene order of HOTAIRM1 $\rightarrow$
HOXA11-AS1 $\rightarrow$ HOTTIP. In the birds, where two or more of
the HOX cluster lncRNA RNA-domains were predicted on the same
scaffold, this gene order was preserved.
%egrep 'HOXA11-AS1|HOTAIR|HOTTIP' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'PCA3' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'
%egrep 'RMST' clans_competed/*gff | perl -lane 'if(/\/(\S+?)\-.*gff:(\S+).*\-id=(\S+);eval/ or /\/(\S+?)\-.*gff:(\S+).*Alias=(\S+);Not/){print "$1\t$2\t$3\t$F[6]"}'

@@

%(neighbor of BRCA1 gene 2)
%MHM results!!!!
%% Another of the functions associated with lncRNAs is chromosomal dosage
%% compensation. In mammals this is carried out by the lncRNAs Xist and
%% Tsix \cite{}. In birds it has been proposed that the lncRNA MHM might
%% play a related role \cite{}.

@@

%% \subsection*{Pseudogenes}

%% The human genome is littered with pseudogenised copies of ncRNAs. The
%% Dfam database of human repeats estimates that approximately $54\%$ of
%% the human genome is composed of repeat elements
%% \cite{Wheeler:2013}. Of these annotations XXX\% are derived from
%% ncRNAs, including the large Alu family of repeats, which is derived
%% from a retransposed SRP RNA pseudogene \cite{}.

%% tRNA, RSV RNA,
%%  \cite{International_Chicken_Genome_Sequencing_Consortium:2004}

%% \emph{Endogenous retroviruses:} RSV RNA, Alfamo CPB

@@

An additonal snoRNA homology search was performed with snoStrip
\cite{Bartschat:2013}. As initial queries we used deutorostomian
snoRNA families from human \cite{Lestrade:2006}, platypus
\cite{Schmitz:2008}, and chicken \cite{Shao:2009}.

SnoRNA annotations overlapping miRNAs annotations where manually
inspected \ppg{and?...}.

@@

Erich Jarvis, Guojie Zhang and Tom Gilbert. Magnus Alm Rosenblad. 
%%%%%%%%%%%%%%%%%%%%%%
%% The Bibliography %%
%%
%% ---------------------------------
%% BioMedCentral bibtex .BST file will be used to
%% create a .BBL file which includes the BMC XML.
%% Note that the displayed Bibliography will not be
%% exactly as specified in the online Instructions for Authors

@@

%% \begin{figure}[ht]
%% \includegraphics[width=0.65\textwidth]{figures/snoRNA-human-yeast-correspondences.pdf}
%% \caption[]{.}\label{fig:2}
%% \end{figure}
\begin{figure}[ht]
  \includegraphics[height=0.95\textheight]{figures/miRNA.pdf}
  \includegraphics[height=0.95\textheight]{figures/snoRNA.pdf}
  \caption[]{Heatmaps showing the prescence/abscence and approximate
    copy number of {\bf miRNA} families on the right and {\bf snoRNA} families on
    the left.}\label{fig:3}
\end{figure}
